Incremental λ-Calculus in Cache-Transfer Style

### 2017-07-12 popl18/static-differentiation.tex
target-language.tex: style revisions

Even though the first two cases of the original transformation are
easily mapped into the two cases of our variant, one may ask where the
third case is realized now. Actually, this third case occurs while we
transform the initial environment. Indeed, we will assume that the
closures of the environment of the source program have been adjoined a
derivative. More formally, we suppose that the derivative of $\sterm$
is evaluated under an environment~$\iderive\sectx$ obtained as
follows:
\[
  \begin{array}{rcl}
    \iderive{\bullet} & = & \bullet \\
    \iderive{\sectxcons\sectx{\tf = \sclosure{\sectx_f}{\lambda \tx. \sterm}}}
    &=&
    \sectxcons{\iderive\sectx}{\tf = \sclosure{\sectx_f}{\lambda \tx. \sterm},
    \tdf = \sclosure{\iderive{\sectx_f}}{\lambda \tx\,\tdx. \iderive{\sterm}}}
    \\
    \iderive{\sectxcons\sectx{\tx = \sclosedvalue}}
    &=&
    \sectxcons{\iderive\sectx}{\tx = \sclosedvalue, \tdx = \inil}
    \qquad \textrm{(If $\sclosedvalue$ is not a closure.)}
  \end{array}
\]

>>>>

Even though the first two cases of \citeauthor{CaiEtAl2014ILC}'s differentiation
map into the two cases of our differentiation variant, one may ask where the
third case is realized now. Actually, this third case occurs while we
transform the initial environment. Indeed, we will assume that the
closures of the environment of the source program have been adjoined a
derivative. More formally, we suppose that the derivative of $\sterm$
is evaluated under an environment~$\iderive\sectx$ obtained as
follows:
\[
  \begin{array}{rcl}
    \iderive{\bullet} & = & \bullet \\
    \iderive{\sectxcons\sectx{\tf = \sclosure{\sectx_f}{\lambda \tx. \sterm}}}
    &=&
    \sectxcons{\iderive\sectx}{\tf = \sclosure{\sectx_f}{\lambda \tx. \sterm},
    \tdf = \sclosure{\iderive{\sectx_f}}{\lambda \tx\,\tdx. \iderive{\sterm}}}
    \\
    \iderive{\sectxcons\sectx{\tx = \sclosedvalue}}
    &=&
    \sectxcons{\iderive\sectx}{\tx = \sclosedvalue, \tdx = \inil}
    \qquad \textrm{(If $\sclosedvalue$ is not a closure.)}
  \end{array}
\]

### 2017-07-12 popl18/target-language.tex
cts-case-studies: abstract over benchmark file location
POPL18: stop saying static caching there
target-language.tex: style revisions

In this section, we present the target language of a transformation
that extends static differentiation with a static caching mechanism.
As said earlier, the functions of $\target$ not only compute their
output but they also returns a cache containing the intermediate
values that have contributed to that output. This cache is transmitted
to derivatives to build changes on top of intermediate values with no
recomputation and derivatives update the caches according to their
input changes.

@@

\input{target-definition}

@@

There are five kinds of change values: closure changes, tuple changes,
literal changes, nil changes and replacements.  Closure changes embed
an environment $\tdectx$ as well as a code pointer for a function
waiting for both a base value $\tx$ and a cache $\tcache$. Notice that
we abusively reuse the same syntax~$\tcache$ to deconstruct and to
construct caches. There other changes are similar to the ones found
in~$\source$.

The base terms of the language are evaluated using a big-step
semantics defined in Figure~\ref{fig:target-definition-base-semantics}.
The judgment~``$\sneval{\tectx}{\tterm}{(\tclosedvalue,
  \tcachedclosedvalues)}$'' is read~``Under the evaluation environment
$\tectx$, the base term $\tterm$ evaluates into a value
$\tclosedvalue$ and a cache $\tcachedclosedvalues$''. The auxiliary
judgment ``$\sneval{\tectx}{\tcache}{\tcachedclosedvalues}$''
specifies the construction of caches.
%
The rule~\rulename{TResult} not only looks into the environment
for the return value~$\tclosedvalue$ but it also evaluates
the returned cache~$\tcache$.
%
The rule~\rulename{TTuple} is similar to the rule of the source
language since no cache is produced by the allocation of a tuple.
%
The rule~\rulename{TClosureCall} works exactly
as~\rulename{SClosureCall} except that the cache value returned by the
closure is bound to the cache identifier $\tcacheid\ty\tf\tx$. The
same remark applies to~\rulename{TPrimitiveCall} with respect
to~\rulename{SPrimitiveCall}.
%
The rule~\rulename{TEmptyCache} evaluates an empty cache term into an
empty cache value. The rule~\rulename{TCacheVar} computes the value of
the cache term~$\tcachecons\tcache\tx$ by appending the value of the
variable~$\tx$ to the cache value~$\tcachedclosedvalues$ computed for
the cache term~$\tcache$. Similarly, the
rule~\rulename{TCacheSubCache} appends the cache value of a cache
named~$\tcacheid\ty\tf\tx$ to the cache value~$\tcachedclosedvalues$
computed for~$\tcache$.

The change terms of the target language are also evaluated using a
big-step semantics defined in
Figure~\ref{fig:target-definition-change-semantics}. The
judgment~``$\sneval{\tdectx}{\tdterm}{(\tcloseddvalue,
  \tcachedclosedvalues)}$'' is read~``Under the evaluation
environment~$\tectx$, the change term $\tdterm$ evaluates into a
change value $\tcloseddvalue$ and an updated cache
$\tcachedclosedvalues$.''. The first auxiliary
judgment~``$\sneval{\tdectx}{\tupdcache}{\tcachedclosedvalues}$''
defines the evaluation of cache update terms. We omit the
rules for this judgment since it is similar to the one for cache terms,
except that cached values are computed by~$\tx \oplus \tdx$, not simply~$\tx$.
%
The final auxiliary
judgment~``$\matchcache\tcachedclosedvalues\tcache\tdectx$'' describes
how a cache pattern $\tcache$ binds a cache
value~$\tcachedclosedvalues$ to produce a change
environment~$\tdectx$.

The rule~\rulename{TDResult} returns the final change value of a
computation as well as a updated cache resulting from the evaluation
of the cache update term~$\tupdcache$.
%
The rule~\rulename{TDTuple} is similar to its counterpart in the
source language, except that no tuple is built for~$\ty$ as it has
already been pushed in the environment by the cache.

>>>>

In this section, we present the target language of a transformation
that extends static differentiation with CTS conversion.
As said earlier, the functions of $\target$ not only compute their
output but they also returns a cache containing the intermediate
values that have contributed to that output. This cache is transmitted
to derivatives to build changes on top of intermediate values with no
recomputation and derivatives update the caches according to their
input changes.

@@

\input{\poplPath{target-definition}}

@@

There are five kinds of change values: closure changes, tuple changes,
literal changes, nil changes and replacements.  Closure changes embed
an environment $\tdectx$ and a code pointer for a function,
waiting for both a base value $\tx$ and a cache $\tcache$. By abuse of notation,
we reuse the same syntax~$\tcache$ to both deconstruct and
construct caches. Other changes are similar to the ones found
in~$\source$.

Base terms of the language are evaluated using a big-step
semantics defined in Figure~\ref{fig:target-definition-base-semantics}.
Judgment~``$\sneval{\tectx}{\tterm}{(\tclosedvalue,
  \tcachedclosedvalues)}$'' is read~``Under evaluation environment
$\tectx$, base term $\tterm$ evaluates to value
$\tclosedvalue$ and cache $\tcachedclosedvalues$''. Auxiliary
judgment ``$\sneval{\tectx}{\tcache}{\tcachedclosedvalues}$''
evaluates cache terms into cache values.
%
Rule~\rulename{TResult} not only looks into the environment
for the return value~$\tclosedvalue$ but it also evaluates
the returned cache~$\tcache$.
%
Rule~\rulename{TTuple} is similar to the rule of the source
language since no cache is produced by the allocation of a tuple.
%
Rule~\rulename{TClosureCall} works exactly
as~\rulename{SClosureCall} except that the cache value returned by the
closure is bound to cache identifier $\tcacheid\ty\tf\tx$.
In the same way, \rulename{TPrimitiveCall} resembles \rulename{SPrimitiveCall}
but also binds $\tcacheid\ty\tf\tx$.
%
Rule~\rulename{TEmptyCache} evaluates an empty cache term into an
empty cache value. Rule~\rulename{TCacheVar} computes the value of
cache term~$\tcachecons\tcache\tx$ by appending the value of
variable~$\tx$ to cache value~$\tcachedclosedvalues$ computed for
cache term~$\tcache$. Similarly,
rule~\rulename{TCacheSubCache} appends the cache value of a cache
named~$\tcacheid\ty\tf\tx$ to the cache value~$\tcachedclosedvalues$
computed for~$\tcache$.

Change terms of the target language are also evaluated using a
big-step semantics, defined in
\cref{fig:target-definition-change-semantics}.
Judgment~``$\sneval{\tdectx}{\tdterm}{(\tcloseddvalue,
  \tcachedclosedvalues)}$'' is read~``Under evaluation
environment~$\tectx$, change term $\tdterm$ evaluates to
change value $\tcloseddvalue$ and updated cache
$\tcachedclosedvalues$''. The first auxiliary
judgment~``$\sneval{\tdectx}{\tupdcache}{\tcachedclosedvalues}$''
defines evaluation of cache update terms. We omit the
rules for this judgment since it is similar to the one for cache terms,
except that cached values are computed by~$\tx \oplus \tdx$, not simply~$\tx$.
%
The final auxiliary
judgment~``$\matchcache\tcachedclosedvalues\tcache\tdectx$'' describes a limited
form of pattern matching used by CTS derivatives: namely,
how a cache pattern $\tcache$ matches a cache
value~$\tcachedclosedvalues$ to produce a change
environment~$\tdectx$.

Rule~\rulename{TDResult} returns the final change value of a
computation as well as a updated cache resulting from the evaluation
of the cache update term~$\tupdcache$.
%
The rule~\rulename{TDTuple} is similar to its counterpart in the
source language, except that no tuple is built for~$\ty$ as it has
already been pushed in the environment by the cache.
